---
title: 'Lab 1: Conceptual foundations'
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/learnr-theme.css
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message = FALSE, warning = FALSE}
## --- Options ---
eval_ex <- TRUE
eval_sol <- TRUE
eval_learnr <- TRUE

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

## --- learnr ---
if ("learnr" %in% (.packages()))
  detach(package:learnr, unload = TRUE)
library(learnr)
```

```{r, echo=FALSE, context="server"}
## --- Options ---
eval_ex <- TRUE
eval_sol <- TRUE
eval_learnr <- TRUE
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## ---- CRAN Packages ----
## Save package names as a vector of strings
pkgs <-  c("foreign", "MASS", "coda", "dplyr", "ggplot2", "ggpubr", "markdown")

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages,
       repos='http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)

## ---- GitHub Packages ----


## ---- Global learnr Objects ----
## Data
gles <- 
  foreign::read.dta("https://github.com/denis-cohen/statmodeling/raw/main/data/gles.dta")

y <- gles$sup_afd
X <- model.matrix( ~ se_self +
                     la_self,
                   data = gles)

## Functions
lm_gibbs <- function(m_scalar,
                     p_scalar,
                     a,
                     b,
                     n_warmup,
                     n_draws,
                     y,
                     X,
                     seed = 20210329,
                     keep_warmup = TRUE) {
  # Set seed
  set.seed(seed)
  
  # Length of chain
  len_chain <- n_warmup + n_draws
  
  # Get vector m, matrix P (note: P = V^(-1))
  m <- rep(m_scalar, ncol(X))
  P <- diag(rep(p_scalar, ncol(X)))
  
  # Neither m_star nor V_star are conditioned on a_star, b_star, or tau
  # So we only need to define these once. As we also need the inverse of
  # V_star, we can store it in an object named P_star.
  V_star <- solve(t(X) %*% X + P)
  P_star <- solve(V_star)
  m_star <- as.vector(V_star %*% (t(X) %*% y + P %*% m))
  
  # Data characteristics
  n_data <- length(y)  

  # Initialize containers
  beta <- matrix(NA, len_chain, ncol(X))
  tau <- rep(NA, len_chain)
  
  # Note that neither 
  
  # Run Gibbs sampler
  for (i in seq_len(len_chain)) {
    if (i == 1) {
      ## Iteration 1: Initialize from prior
      a_star <- a
      b_star <- b
    } else {
      ## Iterations 2+: Update a and b
      a_star <- a + n_data / 2
      b_star <- as.vector(b +
                            0.5 * (t(y) %*% y +
                                     t(m) %*% P %*% m  -
                                     t(m_star) %*% P_star %*% m_star))
    }
    
    ## Sample tau
    tau[i] <- rgamma(1, a_star, b_star)

    ## Sample beta
    beta[i, ] <- MASS::mvrnorm(1, m_star, V_star / tau[i])
  }
  
  ## Conditionally discard warmup-draws
  if (!keep_warmup) {
    tau <- tau[(n_warmup + 1):len_chain]
    beta <- beta[(n_warmup + 1):len_chain, ]
  }
  
  ## Name outputs
  colnames(beta) <- paste0("beta[", colnames(X), "]")
  
  ## Return output
  return(list(beta = beta,
              sigma2 = 1 / tau))
}

## Estimates
# Define 4 seeds
seeds <- sample(10001:99999, 4)

# Run the model with different seeds,collapse to matrix, store as mcmc
gibbs_draws <- lapply(seeds,
                                   function(seed) {
                                     as.mcmc(do.call(
                                       cbind,
                                       lm_gibbs(
                                         m_scalar = 0,
                                         p_scalar = .01,
                                         a = 20,
                                         b = 200,
                                         n_warmup = 5000,
                                         n_draws = 10000,
                                         y = y,
                                         X = X,
                                         seed = seed,
                                         keep_warmup = FALSE
                                       )
                                     ))
                                   })

# Store as mcmc.list
gibbs_draws <- mcmc.list(gibbs_draws)

# Function for formatting text-based solutions
fold_sol <- function(text, title = "Solution") {
  html <- markdown::mark(
    text = text,
    options = c("use_pandoc", "fragment_only")
  )

  cat(
    paste0(
      "<details>\n",
      "<summary><strong>", title, "</strong></summary>\n\n",
      html,
      "\n</details>\n"
    )
  )
}
```

## Learning Goals

- apply core concepts from Bayesian and frequentist inference
- interpret intervals, probabilities, and uncertainty correctly
- diagnose basic features of posterior draws and Markov chains
- summarize and discuss Bayesian model results substantively

## Bayesian and frequentist inference: Interpretations

### Credible intervals

You run a Bayesian analysis to estimate an effect of interest. Your goal is to test the hypothesis that the effect of interest is positive.

You calculate the posterior median, the 95% credible interval, and the 99% credible interval.

The values are as follows:

-   Posterior median: $1.00$
-   95% credible interval: $[0.02; 1.98]$
-   99% credible interval: $[-0.16; 2.16]$

Frequentists would interpret a result with analogous confidence intervals as "statistically significant at the 5%, but not at the 1% level".

What statements could a Bayesian make about the posterior probability in support of the hypothesis of a positive effect?

```{r solution-cred-interval, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
There is strong posterior evidence that the effect is positive. The 95% credible interval lies entirely above zero, and the posterior probability that the effect is greater than zero is between approximately 97.5% and 99.5%.

The 99% credible interval includes zero, which means that the posterior probability that the effect is greater than zero is between approximately 97.5% and 99.5%, as there is between 0.5% and 2.5% posterior probability that the effect is negative (i.e., less than zero).
")
```

### $p$-values

Suppose you have a sample of size $N=150$, from which you recover the sample mean of a continuous variable $X$. Your estimate for the sample mean is $\bar{X} = 129$, your estimate for the standard error of the sample mean is $\hat{\sigma}_{\bar{X}} = 15$.

Let your null hypothesis be that the sample mean is equal to 100. Recover the $p$-value for a two-tailed test. What does this number mean?

```{r p-val-exercise, exercise=TRUE, eval=eval_ex}

```

```{r p-val-exercise-solution, eval=eval_sol, echo=eval_sol}
2 * pnorm(-abs((129 - 100)/ 15))
```

*Hint:* You can click the light bulb icon to see the solution.

```{r p-value-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
The p-value is approximately 0.053. This means that, under the null hypothesis that the true mean equals 100, the probability of observing a sample mean as extreme as 129 (or more extreme) is about 5.3%, interpreted as a long-run frequency across repeated samples. Importantly, p-values do not provide the probability of support for an (alternative) hypothesis. Instead, they indicate the long-run proportion of obtaining results at least as extreme as the observed one under repeated sampling, assuming the null hypothesis is true. Since the p-value is slightly above the conventional significance level of α = 0.05, we fail to reject the null hypothesis, although the result is borderline.
")
```

### A comparison of two plots

In the previous lecture, we saw two types of plots that vaguely resemble one another. The first is the plot from our coin flip experiment:

```{r coin-sim2, eval = TRUE, echo = FALSE, fig.align='center', out.width='75%'}
set.seed(20210329)                   ### set seed for replicability
len_pi <- 1001L                      ### number of candidate values for pi
pi <- seq(0, 1, length.out = len_pi) ### candidate values for pi
a <- b <- 5                          ### hyperparameters
n <- 300                             ### num. of coin flips
pi_true <- .8                        ### true parameter
data <- rbinom(n, 1, pi_true)        ### n coin flips
posterior <- matrix(NA, 3L, n)       ### matrix container for posterior

for (i in seq_len(n)) {    
  current_sequence <- data[1:i]      ### sequence up until ith draw
  k <- sum(current_sequence)         ### number of heads in current sequence
  
  ##### Updating
  a_prime <- a + k               
  b_prime <- b + i - k
  
  ### Analytical means and credible intervals
  posterior[1, i] <- a_prime / (a_prime + b_prime)
  posterior[2, i] <- qbeta(0.025, a_prime, b_prime)
  posterior[3, i] <- qbeta(0.975, a_prime, b_prime)
}

## Plot
plot(
  1:n, 1:n,
  type = 'n',
  xlab = "Number of Coin Flips",
  ylab = expression(
    paste("Posterior Means of ",
          pi,
          sep = " ")
  ), 
  ylim = c(0, 1),
  xlim = c(1, n)
)
abline(
  h = c(.5, .8),
  col = "gray80"
)
rect(-.5, qbeta(0.025, 5, 5),
     0.5, qbeta(0.975, 5, 5),
     col = adjustcolor('red', .4),
     border = adjustcolor('red', .2))
segments(-.5, .5,
         0.5, .5,
         col = adjustcolor('red', .9),
         lwd = 1.5)
polygon(
  c(seq_len(n), rev(seq_len(n))),
  c(posterior[2, ], rev(posterior[3, ])),
  col = adjustcolor('blue', .4),
  border = adjustcolor('blue', .2)
)
lines(
  seq_len(n),
  posterior[1, ],
  col = adjustcolor('blue', .9),
  lwd = 1.5
)
```

The second is the *trace plot*, which we encountered in the context of the convergence diagnostics Gibbs sampler.

```{r trace-example, fig.align='center', out.width='75%', fig.keep="first"}
# Trace plot
coda::traceplot(gibbs_draws)
```


Carefully distinguish what the two types of plots show and what they are used for. Discuss how the meaning of the "iterations" shown along the horizontal $x$-axis differs between the two plots.

```{r plots-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
The first plot shows the evolution of the posterior of a parameter (e.g., the probability of heads π) as more data are observed. The x-axis represents the number of coin flips, meaning the same coin is flipped repeatedly, and with each iteration one additional data point is incorporated through Bayesian updating. The line depicts the posterior mean after each new observation, while the shaded area reflects uncertainty around this estimate. Importantly, at each iteration the posterior distribution is derived analytically rather than via MCMC, meaning it is calculated exactly based on the accumulated data.

The trace plot is used to assess convergence within and across Markov chains, that is, whether the algorithm is correctly sampling from a fixed posterior distribution. The x-axis represents the iterations of the sampler, where each step corresponds to one draw in the MCMC algorithm. The lines show the sequence of posterior samples for the parameter (e.g., β₀), meaning they are draws from the posterior distribution. Over many iterations, these draws collectively approximate the posterior distribution.
")
```

### Random variables

According to the [English-language Wikipedia](https://en.wikipedia.org/wiki/Random_variable), a *"random variable (also called random quantity, aleatory variable, or stochastic variable) is a mathematical formalization of a quantity or object which depends on random events"*. Its distribution is *"a probability measure on the set of all possible values of the random variable*". We usually think of these distributions in terms of known [probability distributions](https://en.wikipedia.org/wiki/Probability_distribution) (i.e., pmf's or pdf's).

Considering this definition and what you have learned today:

-   Why do frequentists think of data, but not of parameters as random variables?
-   Why do Bayesians think of parameters, but not of data as random variables?

```{r rvs-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
Frequentist analysis aims to infer population parameters, such as the population mean of a variable X. These parameters are assumed to be fixed, true, and unknown constants. In contrast, the data are considered random because each sample represents a stochastic draw from the population. Under the assumption of independent and identically distributed (IID) sampling, data could be repeatedly drawn, leading to variation across samples, while the underlying parameter remains unchanged.

In Bayesian inference, the data are treated as fixed once they are observed, while the parameters are considered random variables. This is because unobserved quantities, such as the mean, are assigned probability distributions and are therefore characterized in terms of distributional properties. The data are not treated as random in the analysis because we condition on the observed data. Instead, the prior distribution is updated using the observed data, resulting in the posterior distribution, which reflects our updated beliefs about the parameter.
")
```

### Confidence intervals

Somebody tries to convince you that a 95% confidence interval contains the true population parameter with 95% probability. Why is this incorrect?

```{r cis-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
This statement is incorrect because a given 95% confidence interval does not contain the true parameter with 95% probability or certainty. The interval either contains the true parameter or it does not, so the probability is either zero or one. Instead, a 95% confidence interval represents the long-run proportion of intervals that would contain the true parameter if the sampling process were repeated many times.
")
```

## Posterior summaries and convergence diagnostics

### Prompt

This exercise presents you with three triplets of Markov Chains. These are preloaded in the code chunk below as `chains1`, `chains2`, and `chains3`.

For each object,

1.  Diagnose the chains for any potential problems using the following tools from the `coda` package:
    1.  Trace plots
    2.  Autocorrelation plots
    3.  Potential scale reduction statistic $\hat{R}$ for within and between chain variance
2.  Interpret your diagnosis: What specific problems of non-convergence, if any, do you find?

```{r convdiag-setup}
len_chains <- 2000L
num_chains <- 3L

# ### Chains 1: Well-behaved
# chains <- lapply(1:3,
#                  rnorm(len_chains, mean = 3, sd = 5)) %>%
#   lapply(as.mcmc) %>%
#   as.mcmc.list()

### Chains 2: Well-behaved, but highly autocorrelated
chains3 <- lapply(1:3, function (x) {
  random <- rnorm(len_chains, mean = 3, sd = 5)
  selector <- rep(c(TRUE, rep(FALSE, 9)), 200L)
  random_keep <- c(random[selector], random[len_chains])
  lapply(1:(length(random_keep) - 1L), function(i) {
    tmp <-
      seq(random_keep[i], random_keep[i + 1], length.out = 11L) + rnorm(11L, 0, 0.5)
    return(tmp[1:10])
  }) %>%
    unlist() %>%
    return()
}) %>%
  lapply(as.mcmc) %>%
  as.mcmc.list()

### Chains 3: Between-within variance due to short warm-up
stop_values <- list(
  c(-17, -24, -9, -17, -7, 3),
  c(3, 10, 12, -7, 3, 3),
  c(23, 36, 27, 23, 7, 3)
)
stop_positions <- list(
  c(200, 400, 200, 300, 900),
  c(100, 300, 300, 700, 600),
  c(700, 200, 100, 500, 500)
)
chains1 <- lapply(1:3, function (x) {
  first <- lapply(1:5, function (y) {
    seq(stop_values[[x]][y], stop_values[[x]][y + 1], length.out = stop_positions[[x]][y]) +
      rnorm(stop_positions[[x]][y], mean = 0, sd = 2)
  }) %>%
    unlist()
  second <- rnorm(len_chains / 2, mean = 3, sd = 2)
  return(c(first, second))
}) %>%
  lapply(as.mcmc) %>%
  as.mcmc.list()

### Chains 4: Stationary, but different
start_vals <- c(-1, 3, 7)
chains2 <- lapply(1:3, function (x)
                 rnorm(len_chains, mean = start_vals[x], sd = 5)) %>%
  lapply(as.mcmc) %>%
  as.mcmc.list()

# ### Chains 5: Two parameters, one funky, otherwise well-behaved
# chains <- lapply(1:3, function (x)
#                  data.frame(
#                    par1 = 1 /rgamma(len_chains, 2, 40),
#                    par2 = sample(c(
#                      rbeta(len_chains, 1, 3),
#                      rbeta(len_chains, 3, 1)
#                    ),
#                    len_chains)
#                    )) %>%
#   lapply(as.mcmc) %>%
#   as.mcmc.list()
```

```{r convdiag, exercise = TRUE, exercise.lines = 10}

```

```{r convdiag-solution, eval=eval_sol, echo=eval_sol}
# Trace plot
coda::traceplot(chains1)

# Autocorrelation plot
coda::autocorr.plot(chains1)

# Gelman-Rubin diagnostic
coda::gelman.diag(chains1)

# proceed analogously for chains2 and chains3
```

```{r conv-written-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
Trace plot: At the beginning, the chains are not stationary and show poor mixing; however, this improves after 2000 iterations. This is the same for chains1, chains2, and chains3. It also illustrates why the early part of the chains is typically discarded as warm-up.


Gelman-Rubin: The Gelman-Rubin statistic (R-hat) varies across chains1, chains2, and chains3. Values above 1.1 typically indicate that chains have not fully converged. This is not surprising given the within-chain and between-chain variance we see in the first two thirds of the chains. We can expect that the result would improve after discarding the first 2000 draws as warm-up.

Autocorrelation: This is where chains1, chains2, and chains3 differ most. chains1 shows high autocorrelation, with values close to 1 that decrease only very slowly. This indicates that each draw strongly depends on previous draws in the Markov chain, meaning it takes very long for the sampler to effectively explore the posterior distribution. In chains 3, autocorrelation fades with higher-order lags, and approaches zero after 10-15 draws. This indicates that the sampler explores the posterior inefficiently, but more efficiently than in chains1. In chains2, there is virtually no autocorrelation.
")
```

## Diagnosing, interpreting, and reporting results from a Gibbs sampler for the linear model

### The linear model

Since this is the first regression model we use in a Bayesian estimation framework, it is worthwhile repeating some fundamentals of statistical modeling. Feel free to skip these explanatory sections if you feel you are sufficiently familiar with these concepts.

### Likelihood

The likelihood gives the *generative model* or *data-generating process* for the outcome, $y$.

The linear model stipulates that the observed outcomes $y_i$ for every unit $i$ can be expressed as realizations from a normal distribution with unit-specific *mean or location parameter* $\mu_i$ and a constant (i.e., general) *variance or scale parameter* $\sigma^2$.

$$y_i \sim \text{N}(\mu_i, \sigma^2) \text{ for all }i = 1,...N$$

or, alternatively,

$$y_i = \mu_i + \epsilon_i  \text{ for all }i = 1,...N \\ \epsilon_i \sim \text{N}(0, \sigma^2)$$ The latter notation makes explicit that each observed $y_i$ can be thought of as a combination of a *systematic component*, $\mu_i$, and a *stochastic error component*, $\epsilon_i$, which follows a zero-mean normal distribution with constant variance $\sigma^2$.

### The systematic component

The systematic component is represented by the mean parameter $\mu_i$. In fact, $\mu_i$ is merely a *transformed parameter*: It is a linear function of unit-specific data $\mathbf{x}_i$ and coefficients $\beta$.

The formula below illustrates this, using the row vector notation $\mathbf{x}_i^{\prime} \beta$ as shorthand for the scalar notation $\beta_1 + \beta_2 x_{i, 2} + ...+\beta_k x_{i,k}$.

$$\mu_i = \underbrace{\mathbf{x}_i^{\prime} \beta}_{= \beta_1 + \beta_2 x_{i, 2} + ...+\beta_k x_{i,k}}  \text{ for all }i = 1,...N$$ 

### Parameters and priors

Parameters are the unknown quantities in our models whose posterior distributions we seek to infer.

In the linear model, all coefficients $\beta$ as well as the variance $\sigma^2$ are model parameters. As a variance parameter, $\sigma^2$ must take on strictly positive values.

In Bayesian data analysis, all parameters must be assigned *priors*. For our Gibbs sampler, we assign uninformative, vague priors for all parameters:

-   Independent normal priors for all coefficients: $\beta_k \sim \text{Normal}(0, 100) \text{ for } k = 1,..., K$
-   An inverse Gamma prior for the variance: $\sigma^2 \sim \text{Gamma}^{-1}(20, 200)$.

Our Gibbs sampler will then update these distributions and eventually sample from the posterior target distributions of the five parameters.

### Data

We model respondents' support for the AfD (`sup_afd`, measured on an 11-point scale ranging from -5 to 5) as a function of respondents' pro-redistribution preferences (`se_self`) and anti-immigration preferences (`la_self`).

Both `se_self` and `la_self` are measured on 11-point scales:

-   `se_self` ranges from values (0) "less taxes and deductions, even if that means less social spending" to (10) "more social spending, even if that means more taxes and deductions".
-   `la_self` ranges from values (0) "facilitate immigration" to (10) "restrict immigration".

The model formula is given by

$$\mathtt{sup\_afd} = \beta_1 + \beta_2 \mathtt{se\_self} + \beta_3 \mathtt{la\_self} + \epsilon$$

### Diagnose the estimates

The code chunks below contain preloaded posterior draws from a Gibbs sampler of the linear model. The object is named `gibbs_draws`.

Diagnose the estimates for signs of non-convergence using (at least) traceplots and the Gelman-Rubin diagnostic.

```{r lm-diag, exercise = TRUE, exercise.lines = 10}

```

```{r lm-diag-solution, eval=eval_sol, echo=eval_sol}
# Trace plot
coda::traceplot(gibbs_draws)

# Gelman-Rubin diagnostic
coda::gelman.diag(gibbs_draws)
```

```{r lm-written-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
None of the four plots show signs of non-convergence. All plots show no trends (no upward and downward drift), appear stationary (the chains fluctuate around a stable value), and there is a strong overlap between the chians. Similarly, the Gelman-Rubin diagnostic remains below 1.1 for all parameters.
")
```

### Summarize the estimates

The code chunk below contains code that pools the posterior draws across all chains. The resulting objects is a data frame, where each column holds the stacked posterior draws of a given parameter.

Summarize the posterior distribution in terms of posterior medians, posterior standard deviations, and the 95% credible interval using `apply()`.

```{r lm-sum, exercise = TRUE, exercise.lines = 10}
# Pool posterior draws across chains
pooled_gibbs_draws <- do.call(rbind.data.frame, gibbs_draws)
```

```{r lm-sum-solution, eval=eval_sol, echo=eval_sol}
# Pool posterior draws across chains
pooled_gibbs_draws <- do.call(rbind.data.frame, gibbs_draws)

# Summarize using apply
pooled_gibbs_draws %>%
  apply(2, function (x) {
    c(
      "Median" = median(x),
      "SD" = sd(x),
      quantile(x, c(.025, .975))
    )
  }) %>%
  round(2)
```

What are your substantive conclusions?

```{r lm-subst-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
The posterior median represents the central tendency, the standard deviation describes the spread of the effect, and the interval provides a quantile-based summary.
These summaries therefore describe the posterior distribution of each regression parameter analogously to how point estimates, standard errors, and confidence intervals summarize sampling distributions in frequentist analysis.

An interpretation example is the slope paramter for anti-immigration preferences, beta[la_self]. We learn that a one-unit increase toward more restrictive immigration preferences on the 0-10 scale predicts a 0.36 point increase in AfD support, measured on a -5 to +5 scale. The 95% credible intervals ranges from 0.31 and 0.41, which indicates a very high probability for a sizable positive effect.
")
```

## Probability distributions

### Probability distributions in statistical modeling and inference

In statistical modeling and statistical inference, we commonly encounter probability distributions in two distinct roles:

1.  As distributional characterizations of data-generating processes of outcome variables (e.g., "each single flip of a fair coin is a draw from a Bernoulli distribution with probability parameter 0.5").
2.  As distributional characterizations of inferential uncertainty about model parameters, either in the form of sampling distributions (in frequentist inference) or in the form of prior and posterior distributions (in Bayesian inference).

### Matching distributions and outcomes

To get familiar with some common distributions, briefly skim the top info boxes in the linked Wikipedia articles. They typically convey some central information, such as:

-   The conventional notation of the distribution
-   Its parameters (and their constraints)
-   The support of the distribution
-   Mathematical formulas for the pdf/pmf, CDF, and moments (mean, variance, etc)
-   Visualizations of prototypical distributions given a selection of parameters

1.  [Pareto distribution](https://en.wikipedia.org/wiki/Pareto_distribution)
2.  [Bernoulli distribution](https://en.wikipedia.org/wiki/Bernoulli_distribution)
3.  [Weibull distribution](https://en.wikipedia.org/wiki/Weibull_distribution)
4.  [Beta distribution](https://en.wikipedia.org/wiki/Beta_distribution)
5.  [Categorical distribution](https://en.wikipedia.org/wiki/Categorical_distribution)
6.  [Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution)

Which of these distribution would you most likely use to model the following outcomes?

a.  Individuals' time spent in unemployment before reemployment
b.  Countries' unemployment rates
c.  Voters' vote choices in a multi-party election
d.  Countries' income distributions
e.  Individuals' decisions to turnout in an election
f.  Parliamentarians' number of missed plenary sessions

```{r dists1-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
a.  Individuals' time spent in unemployment before reemployment: Weilbull
b.  Countries' unemployment rates: Beta
c.  Voters' vote choices in a multi-party election: Categorical  
d.  Countries' income distributions: Pareto 
e.  Individuals' decisions to turnout in an election: Bernoulli 
f.  Parliamentarians' number of missed plenary sessions: Poisson
")
```

### Bivariate density distributions

Statistical inference often involves the estimation of joint distribution of multiple model parameters. The below shows a hypothetical joint distribution of two correlated parameters.

-   The marginal distributions are characterized by two distributions we encountered as part of the lecture.
    -   Can you guess what these distributions are?
    -   Can you make any approximate statements about the parameters that characterize these distributions?
-   What can you say about the sign of the correlation between the two?

```{r prior-plot, fig.align='center', fig.width = 6, fig.height = 6}
jd <- tibble::tibble(
  x = rbeta(10000L, 0.75, 0.75)
) %>%
  dplyr::mutate(
    y = rnorm(10000L, (x - 0.5) * 4, sd = 2.5)
  )

## Joint density
joint <- jd %>%
  ggplot(aes(x = x, y = y)) +
  stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
  scale_fill_viridis_c() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  xlab("x") +
  ylab("y")   +
  theme(legend.position = "none") +
  theme(plot.margin = unit(c(0, 0, 1, 1), 'lines'))

hist_x <- jd %>%
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  theme(plot.margin = unit(c(0, 0, 0, 0), 'lines'))

hist_y <- jd %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  rotate() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  theme(plot.margin = unit(c(0, 0, 0, 0), 'lines'))

ggarrange(
  hist_x,
  NULL,
  joint,
  hist_y,
  nrow = 2,
  ncol = 2,
  align = "hv",
  widths = c(2, 1),
  heights = c(1, 2)
)
```

```{r dists2-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
Judging by the shape of the univariates marginal distributions, we can infer that x resembles a Beta distribution, y resembles a normal distribution.

Judging by their respective distributional patterns, we can infer that x has a symmetrical bimodal shape. This shows that the rate and shape parameters are equal, and below one. We see that y is centered at zero and that most of the probability density lies between -5 and +5; this shows that it is a normal distribution with mean 0 and standard deviation 2.5.

Lastly, judging by the bivariate density, we see that higher levels of x tend to go hand in hand with higher levels of y (and vice versa). This shows that both distributions are positively correlated.
")
```
