---
title: 'Lab 2: Using brms'
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/learnr-theme.css
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
## --- Options ---
eval_ex <- TRUE
eval_sol <- TRUE
eval_learnr <- TRUE

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

## --- learnr ---
if ("learnr" %in% (.packages()))
  detach(package:learnr, unload = TRUE)
library(learnr)
```

```{r, echo=FALSE, context="server"}
## --- Options ---
eval_ex <- TRUE
eval_sol <- TRUE
eval_learnr <- TRUE
```

```{r setup, include=FALSE}
## ---- CRAN Packages ----
## Save package names as a vector of strings
pkgs <-
  c("rstan",
    "brms",
    "future",
    "bayesplot",
    "dplyr")

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())],
       install.packages,
       repos = 'http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)

## ---- Data ----
lab3_url <-
  "https://github.com/denis-cohen/statmodeling/raw/main/data/lab3.RData"
load(url(lab3_url))

# Function for formatting text-based solutions
fold_sol <- function(text, title = "Solution") {
  html <- markdown::mark(
    text = text,
    options = c("use_pandoc", "fragment_only")
  )

  cat(
    paste0(
      "<details>\n",
      "<summary><strong>", title, "</strong></summary>\n\n",
      html,
      "\n</details>\n"
    )
  )
}
```

## Learning Goals

- fit Bayesian models in brms for applied research questions
- interpret posterior estimates and uncertainty substantively
- assess model diagnostics and model fit
- communicate results clearly and appropriately

## Logit model

### The three parts of every GLM

All generalized linear models have three characteristic parts:

#### Family

-   The family stipulates a stochastic process that can plausibly generate an outcome $y$
-   This means we choose a pdf or pmf for $y$ given some parameters: $y_i \sim \text{f}(\theta_i, \psi)$
-   The choice usually depends on the distributional properties of $y$

#### Linear component

-   A linear model $y_i^{\ast} = \mathbf{x}_i^{\prime} \beta + \epsilon_i$
-   The goal of inference is the estimation of $\beta$
-   From, this, we can derive our *systematic component* or *linear predictor*, $\eta_i = \mathbf{x}_i^{\prime} \beta$

#### Inverse link function

-   A function that transforms the systematic component $\eta_i$ such that it represents a characteristic *parameter* $\theta_i$ of the family
-   $\theta_i = g^{-1}(\eta_i)$

### The logit model

The logit model is a popular model for binary outcomes.

-   Family: $$y_i \sim \text{Bernoulli}(\pi_i)$$
-   Linear component: $$y_i^{\ast} = \underbrace{\mathbf{x}_i^{\prime} \beta}_{\eta_i} + \underbrace{\epsilon_i}_{\sim \text{Logistic}(0, 1)}$$
-   Inverse link function: $$\pi_i = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}$$

Thus, the logit model is given by

$$y_i \sim \text{Bernoulli}\left(\frac{\exp(\eta_i)}{1 + \exp(\eta_i)}\right)$$

where the $\beta$ vector is being estimated. Note that the parameters of the $\text{Logistic}(0, 1)$ error distribution are fixed and therefore need not be estimated.

### Context

In this exercise, we are interested in how the effect of socio-economic policy preferences and immigration policy preferences affected vote choices for the Greens in 2017.

The following are once again or focal predictors:

-   `se_self` ranges from values (0) "less taxes and deductions, even if that means less social spending" to (10) "more social spending, even if that means more taxes and deductions".
-   `la_self` ranges from values (0) "facilitate immigration" to (10) "restrict immigration".

### Data

We model respondents' vote choices for the Greens (`vote_greens`), as a function of respondents' pro-redistribution preferences (`se_self`) and anti-immigration preferences (`la_self`), each of which we interact with respondents' gender. We also control for age (`age`) and East/West residence (`east`).

The model formula is given by

$$
\mathtt{vote\_greens}_i = \\\beta_1 + \beta_2 \mathtt{la\_self}_i + \beta_3 \mathtt{se\_self}_i + \\ \beta_4 \mathtt{fem}_i + \beta_5 \mathtt{east}_i + \beta_6 \mathtt{age}_i + \\ \beta_7 \mathtt{la\_self}_i \times \mathtt{fem}_i + \beta_8 \mathtt{se\_self}_i \times \mathtt{fem}_i + \epsilon
$$

## Fitting

### Choosing priors

Check the default priors for the logit model using `brms::get_prior()`.

Note that you must supply the model formula, the `data`, and the `family` of the model. For a logit model, we must set `family = bernoulli(link = "logit")`

```{r brms-model-1, exercise=TRUE, eval=eval_ex}

```

```{r brms-model-1-solution, eval=eval_sol, echo=eval_sol}
# Get default priors
default_priors_logit <- brms::get_prior(
  vote_greens ~                          # outcome
    la_self +                            # immigration preferences
    se_self +                            # redistribution preferences
    fem +                                # gender
    east +                               # east/west residence
    age +                                # age
    la_self:fem +                        # interaction term 1
    se_self:fem,    
  data = gles,                          # data
  family = bernoulli(link = "logit")    # family and link
)
default_priors_logit
```

*Note:* Missing entries in the `prior` column denote flat/uniform priors.

### Define custom priors

Generate an object called `custom_priors_logit` with $\text{Normal(0,5)}$ priors for both intercept and slopes. 

```{r brms-model-2, exercise=TRUE, eval=eval_ex, exercise.lines = 6}

```

```{r brms-model-2-solution, eval=eval_sol, echo=eval_sol}
custom_priors_logit <- c(
  brms::prior(normal(0, 5), class = b),          # normal slopes
  brms::prior(normal(0, 5), class = Intercept)   # normal intercept
)
custom_priors_logit
```

### Fitting the model

Fit the model using `brms::brm()`. Pass your custom priors from `custom_priors_logit` to the model. 

Run the model with four chains for 2000 iterations each, discarding the first 1000 draws as warmup. Set the seed to today's date, `20260417L`.

*Note:* Model compilation and estimation may take a while.

```{r brms-model-3, exercise=TRUE, eval=eval_ex, exercise.lines = 18, exercise.timelimit = 180}

```

```{r brms-model-3-solution, eval=eval_sol, echo=eval_sol}
logit_brms <- brms::brm(
  vote_greens ~                          # outcome
    la_self +                            # immigration preferences
    se_self +                            # redistribution preferences
    fem +                                # gender
    east +                               # east/west residence
    age +                                # age
    la_self:fem +                        # interaction term 1
    se_self:fem,                         # interaction term 2 
  data = gles,                           # data
  family = bernoulli(link = "logit"),    # family and link
  prior = custom_priors_logit,           # priors
  chains = 4L,                           # number of chains
  iter = 2000L,                          # number of iterations per chain
  warmup = 1000L,                        # number of warm-up samples per chain
  algorithm = "sampling",                # algorithm (HMC/NUTS)
  backend = "rstan",                     # backend (rstan)
  seed = 20260417L                       # seed
)
```

## Summarize and diagnose

### Model summary and generic diagnostics

Print the model summary and check `Rhat` for any signs of non-convergence.

```{r brms-print, exercise=TRUE, eval=eval_ex, exercise.lines = 2}

```

```{r brms-print-solution, eval=eval_sol, echo=eval_sol}
logit_brms
```

```{r rhat-written-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
For all coefficients, the Rhat is below 1.1, showing that there are no signs of non-convergence. 
")
```

### Visual diagnostics

Explore some visual diagnostics of your choice. See `help(mcmc_plot)` for types of plots.

```{r brms-visual, exercise=TRUE, eval=eval_ex, exercise.lines = 5}

```

```{r brms-visual-solution, eval=eval_sol, echo=eval_sol}
brms::mcmc_plot(logit_brms, type = "rhat")  # Gelman-Rubin 
brms::mcmc_plot(logit_brms, type = "acf")   # Autocorrelation
brms::mcmc_plot(logit_brms, type = "trace") # Trace plots
```

```{r diag-plots-written-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
Gelman-Rubin: All Rhat values are below 1.05, indicating that the chains have converged well. 

Autocorrelation: For all parameters across all chains, autocorrelation decreases rapidly and approaches zero after only a few lags, indicating efficient mixing.

Trace plots: The trace plots indicate good convergence of the MCMC chains. First, all chains are stationary, as they fluctuate around a constant level without any visible upward or downward trends. Second, the chains exhibit good mixing, moving rapidly up and down and forming a 'fuzzy band', which suggests low autocorrelation. Third, the chains overlap closely for each parameter and are largely indistinguishable from one another, indicating that they are sampling from the same posterior distribution.
")
```

### Algorithm-specific diagnostics

Use the `check_hmc_diagnostics()` function from the `rstan` package to check for algorithm-specific anomalies. Remember that we must extract the `stanfit` object nested in our `brmsfit` object via `logit_brms$fit`.

```{r algo-diag, exercise=TRUE, eval=eval_ex, exercise.lines = 2}

```

```{r algo-diag-solution, eval=eval_sol, echo=eval_sol, message=TRUE, warning=TRUE}
rstan::check_hmc_diagnostics(logit_brms$fit)
```

## Quantities of interest

Use `marginaleffects::plot_slopes()` to visualize the effects of (1) immigration preferences (`la_self`) and (2) socio-economic preferences (`se_self`), each conditional on respondents' gender (`fem`), on the probability of voting for the Greens.

Interpret your findings.

```{r mfx-la-se, exercise=TRUE, eval=eval_ex, exercise.lines = 12}

```

```{r mfx-la-se-solution, eval=eval_sol, echo=eval_sol}
marginaleffects::plot_slopes(logit_brms,
                             variable = "la_self",
                             condition = "fem") +
  ggplot2::geom_hline(yintercept = 0,
                     col = "red",
                     alpha = 0.5) +
  ggplot2::xlab("Gender (Men = 0, Women = 1)") +
  ggplot2::ylab(paste(
    "Marginal effect of right-wing immigration",
    "policy preferences on Greens support",
    sep = "\n"
  ))

marginaleffects::plot_slopes(logit_brms,
                             variable = "se_self",
                             condition = "fem") +
  ggplot2::geom_hline(yintercept = 0,
                     col = "red",
                     alpha = 0.5) +
  ggplot2::xlab("Gender (Men = 0, Women = 1)") +
  ggplot2::ylab(
    paste(
      "Marginal effect of left-wing socio-economic",
      "policy preferences on Greens support",
      sep = "\n"
    )
  )
```

```{r subst1-written-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
The marginal effect of right-wing immigration preferences on Green support is negative for both men and women. The effect appears to be more negative for women than for men. One may wonder if the magnitude of the negative effect among women is greater than that of the effect among men; to determine this, we would have to analyze the posterior of the effect difference.

The marginal effect of left-wing socio-economic policy preferences on Green support is positive for men, as the credible interval lies entirely above zero. For women, the credible interval includes zero and negative values, indicating greater uncertainty about the direction of the effect.
")
```

## Posterior predictive checks

### Distributional congruence

Use posterior predictive checks to test for distributional congruence.

Note that since our outcome is binary, you will learn more from using `type = "bars"` than `type = "dens_overlay"`.

```{r brms-lm-pp-1, exercise=TRUE, eval=eval_ex, exercise.lines = 3}

```

```{r brms-lm-pp-1-solution, eval=eval_sol, echo=eval_sol}
brms::pp_check(logit_brms, ndraws = 1000, type = "bars")
```

What do you conclude? Does our model get the numbers of green voters right?

```{r pp1-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
Yes, the model reliably reproduces the aggregate distribution of our outcome variable: the posterior predictions closely resemble the distributional shape of the observed data.
")
```

### Observation-level prediction error

Now turn to the observation-level prediction error. This is best assessed by setting the arguments `type = "error_hist_grouped"` and `group = "vote_greens"`. What can you conclude from this plot?

```{r brms-lm-pp-2, exercise=TRUE, eval=eval_ex, exercise.lines = 3}

```

```{r brms-lm-pp-2-solution, eval=eval_sol, echo=eval_sol}
brms::pp_check(logit_brms, ndraws = 1, type = "error_hist_grouped", group = "vote_greens")
```

```{r pp2-solution, echo=FALSE, eval=eval_sol, results='asis'}
fold_sol("
The model struggles to accurately identify respondents as Green voters. Many actual Green voters are predicted as non-Green voters, and vice versa.
")
```
