---
title: "Bayesian Modeling"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/learnr-theme.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
if ("learnr" %in% (.packages()))
  detach(package:learnr, unload = TRUE)
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

## Save package names as a vector of strings
pkgs <-  c()

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages,
       repos='http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)
```

## Introductions

### Hi, I'm Denis

- Senior Research Fellow, Data and Methods Unit, MZES
- Dr. phil. (political science), Humboldt University of Berlin (2018)
- I work on:
    - political consequences of spatial inequalities and residential mobility
    - party competition
    - voting behavior
    - quantitative social science (both pedagogical and methodological)

### Hi, I'm Maria
 
- Master's Student in Sociology, University of Mannheim
- B.A. (Political Science and Economics), Goethe University Frankfurt am Main (2024)
- Student Assistant at Mannheim Centre for European Social Research
- I'm interested in: political sociology, quantitative methods 

### Who are you?

- Your name
- Your discipline and current position
- What brought you here

## Bayesian Modeling: From Foundations to Custom Solutions

###  Week 1: Learning Objectives

#### Conceptual: Foundations of Bayesian inference

1. Contrast frequentist and Bayesian inference paradigms
1. Understand when and why researchers go Bayesian
1. Understand the basics of analytical and (especially) numerical Bayesian inference
1. Understand the costs and benefits of doing applied Bayesian data analysis

#### Practical: Applied Bayesian data analysis using pre-implemented tools

1. Estimate and diagnose models using `brms`
1. Process and report model estimates using `modelsummary` and `marginaleffects`

### Day 1: Bayesian and frequentist inference: A juxtaposition

- Conceptual foundations and intuition
- Little to no code and programming
- Focus on *understanding*

### Day 2: Applied Bayesian statistics: A gentle introduction

- Applied data analysis using pre-implemented tools
- Estimating and processing Bayesian models via `brms`, `modelsummary`, and `marginaleffects`
- Focus on *using* existing tools 

###  Week 2: Learning Objectives

#### Conceptual: A peek under the hood

1. Understand generalized linear models as a common theoretical framework for statistical models
1. Understand how to process estimates from generalized linear models into substantively meaningful quantities of interest *without* relying on pre-implemented software

#### Practical: Custom model building in Stan

1. Learn how to translate your own statistical models into code using Stan
1. Learn how to implement the Bayesian workflow as a model developer, as opposed to a model user

### Self-learning: Take-home exercise

As a segue from Week 1 to Week 2, I have prepared a self-learning module that covers the following topics:

- Linear algebra and programming basics (self-learn module)
- Applied statistical programming in R

### Self-learning: Your own project

- Becoming an active Bayesian practitioner requires practice!
- The challenges you may encounter along the way will likely be idiosyncratic

Therefore, I encourage you to:

  - Start exploring Bayesian implementations using `brms` after Week 1
  - Attend individual consultations at the end of Session 4 to address any questions you have encountered

*Note:* At the same time, Maria will be available for consultations about any aspects of the course.


### Day 3: Statistical programming in R and Stan: An advanced introduction

- Generalized linear models and quantities of interest
- Coding and validating a Bayesian model from scratch using Stan
- Focus on *building* your own models

### Day 4: Applied Bayesian modeling in Stan

- Extensions to the basic workflow
- Post-processing posterior draws
- Choosing between pre-implemented and custom solutions
- **Individual consultations**: Bring your own data and model and get help with implementation, inference, and workflow

### In short

#### Week 1

1. Understand concepts (today)
1. Use existing tools (tomorrow)

#### Week 2

1. Understand underlying mechanics
1. Extend tools
1. Decide how to proceed in your own work

### A quick note on mathematical notation

- The teaching materials for this course use mathematical notation
- It is helpful, albeit not strictly necessary that you understand it:
    - Mathematical notation is a compact and universal tool to convey concepts...
    - ...but what really matters is that you understand the intuition of the underlying concepts...
    - ...and you will not be asked to translate formulas into code.
- If anything is unclear: Please ask at any time!

## Course materials

### GitHub

All course materials are available on my GitHub: https://github.com/denis-cohen/bayesianmodeling 

### Interactive tutorials

The primary course materials come as `learnr` tutorials wrapped in an R package, `bayesianmodeling`.

You can use the interactive materials to:

- click along during the lectures
- work on your lab exercises

To download, install, and use the interactive materials, run this code that you can copy from GitHub and paste into R:

```{r, eval=FALSE, echo=TRUE}
# Package installation ----
## Detach if loaded ----
if ("bayesianmodeling" %in% (.packages())) {
  detach(package:bayesianmodeling, unload = TRUE)
}

## Uninstall if installed ----
if ("bayesianmodeling" %in% installed.packages()) {
  remove.packages("bayesianmodeling")
}

## Install if not installed ----
if (!("devtools" %in% installed.packages())) {
  install.packages("devtools")
}

## Load from GitHub ----
library(devtools)
devtools::install_github("denis-cohen/bayesianmodeling")

## Load to library ----
library(bayesianmodeling)

# Run the tutorials (one at a time) ----
## Introduction ----
learnr::run_tutorial("00-00-int", package = "bayesianmodeling")

## Day 1 ----
learnr::run_tutorial("01-01-lec", package = "bayesianmodeling")
learnr::run_tutorial("01-02-lec", package = "bayesianmodeling")
learnr::run_tutorial("01-03-lab", package = "bayesianmodeling")

## Day 2 ----
learnr::run_tutorial("02-01-lec", package = "bayesianmodeling")
learnr::run_tutorial("02-02-lec", package = "bayesianmodeling")
learnr::run_tutorial("02-03-lab", package = "bayesianmodeling")

## Take home exercise ----
learnr::run_tutorial("03-00-the", package = "bayesianmodeling")
```

### Static course materials

#### Course notes

I have also rendered the `learnr` tutorials (minus the interactive components) as static documents:

- Lecture notes come in PDF format
- Lab notes come in HTML format in two versions: with and without code-based solutions and outputs 

Along with the data used in this course, these materials are available:

1. from the GESIS Moodle course
1. from my GitHub (see subfolders `lecture-materials`, `lab-materials`, and `data`)

## Schedule

### Structure of our sessions

1. Lecture:
    - Input
    - Dedicated Q&A (but feel free to ask for clarification at any point!)
2. Lab: Hands-on exercises (in small teams)
    - You will be assigned to small teams in breakout rooms
    - I will be in the main meeting, available for questions
    - The TA will switch between breakout rooms to assist you
    - We will jointly discuss your answers and the solutions in the main meeting

### RStan/brms installation: Any problems?

Did you successfully get RStan to run on your machine, incl. `example(stan_model, package = "rstan", run.dontrun = TRUE) `?

- If yes: Great! You're all set.
- If no: No worries, we will start using `brms` *tomorrow*. I will stick around after 15:15 today in case you need assistance.

## Questions?




